---
title: "Cohorts"
format:
  html:
    toc: true
    toc_depth: 3
    html-math-method: katex
    grid:
      body-width: 1200px
---

This Documents takes a look at who are in the cohorts.

- Aim 1 needs people in the cognition substudy who were free of dementia at year 5

- Aim 2 needs people with repeated 3MSE/DSST scoress

- Make a flowchart with the N at each step, for Soma, Olink and Elisa separately.

# Setup

```{r}
#| message: false

rm(list = ls())

library(tidyverse)
library(haven)
library(gt)
library(gtsummary)
library(glue)
library(writexl)
library(readxl)
library(ggVennDiagram)
library(glue)
library(ggpubr)



# Import data
source('paths.R')
```

## Clinical data

```{r}
data_clin <- read_csv(
  file.path(
    data_path, 'Data_processed', 'Meineche_10237_recoded.csv')
  )
```

## Protein data

```{r}
data_soma <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'soma_y5_klotho.xlsx'
  ), 
  sheet = 'values_scaled'
)

idnos_soma <- data_soma %>% pull(idno)

data_olink <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'olink_y5_klotho.xlsx'
  ), 
  sheet = 'values_scaled'
)

idnos_olink <- data_olink %>% pull(idno)

data_elisa <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'elisa_klotho.xlsx'
  ), 
  sheet = 'values_scaled'
)

idnos_elisa <- data_elisa %>% pull(idno)


# rm(data_soma, data_olink)
```

# Participants

## Venn

```{r}
#| fig-width: 6
#| fig-height: 6


ggVennDiagram(list(Olink = idnos_olink, SomaScan = idnos_soma, Elisa =idnos_elisa),
              label_alpha = 0, edge_size = 0.5,
              set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none")
```

## Soma

```{r}

data_soma_clin <- data_clin %>%
  filter(idno %in% idnos_soma)

n_start <- nrow(data_soma_clin)

message(glue('Start: N={n_start}'))

message('AIM 1:')

data_soma_clin <- data_soma_clin %>%
  filter(!is.na(demclass))

n_not_cog <- nrow(data_soma_clin)

message(glue('{n_start-n_not_cog} not in cognition study (no dementia information).'))

data_soma_clin <- data_soma_clin %>%
  filter(demclass != 'Prevalent')

n_prev <- nrow(data_soma_clin)

message(glue('{n_not_cog-n_prev} removed due to prevalent dementia. N={n_prev}.'))

data_soma_clin <- data_soma_clin %>%
  filter(yearsfu > 1)

n_one_year <- nrow(data_soma_clin)

message(glue('{n_prev-n_one_year} removed due dementia within first year. N={n_one_year}.'))

idnos_soma_aim1 <- data_soma_clin %>% pull(idno)


# ------------------------------------------------------------------------------------------

message('\nAIM 2:')

data_soma_clin_aim2 <- data_soma_clin

fu_3mse <- paste0('score134_y', 6:11)
fu_dsst <- paste0('digcor_y', 6:11)

aim2_3mse <- data_soma_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))

aim2_dsst <- data_soma_clin_aim2 %>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))

idnos_soma_aim2_3mse <- aim2_3mse %>% pull(idno)
idnos_soma_aim2_dsst <- aim2_dsst %>% pull(idno)

message(glue('3MSE: {n_one_year - nrow(aim2_3mse)} removed (no Y5 score or no follow-up). N={nrow(aim2_3mse)}'))
message(glue('DSST: {n_one_year - nrow(aim2_dsst)} removed (no Y5 score or no follow-up). N={nrow(aim2_dsst)}'))

data_soma_clin_aim2 <- data_soma_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))%>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))

```

## Olink

```{r}
data_olink_clin <- data_clin %>%
  filter(idno %in% idnos_olink)

n_start <- nrow(data_olink_clin)

message(glue('Start: N={n_start}'))

message('AIM 1:')

data_olink_clin <- data_olink_clin %>%
  filter(!is.na(demclass))

n_not_cog <- nrow(data_olink_clin)

message(glue('{n_start-n_not_cog} not in cognition study (no dementia information).'))


data_olink_clin <- data_olink_clin %>%
  filter(demclass != 'Prevalent')

n_prev <- nrow(data_olink_clin)

message(glue('{n_not_cog-n_prev} removed due to prevalent dementia. N={n_prev}.'))

data_olink_clin <- data_olink_clin %>%
  filter(yearsfu > 1)

n_one_year <- nrow(data_olink_clin)

message(glue('{n_prev-n_one_year} removed due dementia within first year. N={n_one_year}.'))

idnos_olink_aim1 <- data_olink_clin %>% pull(idno)
  
  
# -----------------------------------------------------------------------------------


message('\nAIM 2:')

data_olink_clin_aim2 <- data_olink_clin

fu_3mse <- paste0('score134_y', 6:11)
fu_dsst <- paste0('digcor_y', 6:11)

aim2_3mse <- data_olink_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))

aim2_dsst <- data_olink_clin_aim2 %>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))

idnos_olink_aim2_3mse <- aim2_3mse %>% pull(idno)
idnos_olink_aim2_dsst <- aim2_dsst %>% pull(idno)


message(glue('3MSE: {n_one_year - nrow(aim2_3mse)} removed (no Y5 score or no follow-up). N={nrow(aim2_3mse)}'))
message(glue('DSST: {n_one_year - nrow(aim2_dsst)} removed (no Y5 score or no follow-up). N={nrow(aim2_dsst)}'))

data_olink_clin_aim2 <- data_olink_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))%>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))
```

## Elisa

```{r}
data_elisa_clin <- data_clin %>%
  filter(idno %in% idnos_elisa)

n_start <- nrow(data_elisa_clin)

message(glue('Start: N={n_start}'))

message('AIM 1:')

data_elisa_clin <- data_elisa_clin %>%
  filter(!is.na(demclass))

n_not_cog <- nrow(data_elisa_clin)

message(glue('{n_start-n_not_cog} not in cognition study (no dementia information).'))


data_elisa_clin <- data_elisa_clin %>%
  filter(demclass != 'Prevalent')

n_prev <- nrow(data_elisa_clin)

message(glue('{n_not_cog-n_prev} removed due to prevalent dementia. N={n_prev}.'))

data_elisa_clin <- data_elisa_clin %>%
  filter(yearsfu > 1)

n_one_year <- nrow(data_elisa_clin)

message(glue('{n_prev-n_one_year} removed due dementia within first year. N={n_one_year}.'))

idnos_elisa_aim1 <- data_elisa_clin %>% pull(idno)
  
  
# -----------------------------------------------------------------------------------


message('\nAIM 2:')

data_elisa_clin_aim2 <- data_elisa_clin

fu_3mse <- paste0('score134_y', 6:11)
fu_dsst <- paste0('digcor_y', 6:11)

aim2_3mse <- data_elisa_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))

aim2_dsst <- data_elisa_clin_aim2 %>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))

idnos_elisa_aim2_3mse <- aim2_3mse %>% pull(idno)
idnos_elisa_aim2_dsst <- aim2_dsst %>% pull(idno)


message(glue('3MSE: {n_one_year - nrow(aim2_3mse)} removed (no Y5 score or no follow-up). N={nrow(aim2_3mse)}'))
message(glue('DSST: {n_one_year - nrow(aim2_dsst)} removed (no Y5 score or no follow-up). N={nrow(aim2_dsst)}'))

data_elisa_clin_aim2 <- data_elisa_clin_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))%>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))
```

```{r}
nrow(data_olink_clin)
length(idnos_olink_aim1)
nrow(data_elisa_clin)
length(idnos_elisa_aim1)
```

```{mermaid}
flowchart TD
    subgraph SOMA["<b>SomaScan</b>"]
        direction TB
        S0["CHS Year 5 participants<br/>with SomaScan<br/><b>N = 3,171</b>"]
        SX1["<b>Excluded, n = 875</b><br/>Not in Cognition Study: 789<br/>Prevalent dementia: 86"]
        S1["Dementia-free at baseline<br/><b>N = 2,296</b>"]
        SX2["<b>Excluded, n = 30</b><br/>Dementia onset &lt; 1 year"]
        S2["<b>Aim 1: Incident dementia</b><br/>N = 2,266"]
        SX3["<b>Excluded</b><br/>No Y5 score or follow-up<br/>3MSE: 35 · DSST: 74"]
        S3["<b>Aim 2: 3MSE</b><br/>N = 2,231"]
        S4["<b>Aim 2: DSST</b><br/>N = 2,192"]

        S0 --> S1
        S0 -.- SX1
        S1 --> S2
        S1 -.- SX2
        S2 --> S3
        S2 --> S4
        S2 -.- SX3
    end
    
    subgraph OLINK["<b>Olink</b>"]
        direction TB
        O0["CHS Year 5 participants<br/>with Olink<br/><b>N = 4,037</b>"]
        OX1["<b>Excluded, n = 1,251</b><br/>Not in Cognition Study: 1,088<br/>Prevalent dementia: 163"]
        O1["Dementia-free at baseline<br/><b>N = 2,786</b>"]
        OX2["<b>Excluded, n = 60</b><br/>Dementia onset &lt; 1 year"]
        O2["<b>Aim 1: Incident dementia</b><br/>N = 2,726"]
        OX3["<b>Excluded</b><br/>No Y5 score or follow-up<br/>3MSE: 64 · DSST: 119"]
        O3["<b>Aim 2: 3MSE</b><br/>N = 2,662"]
        O4["<b>Aim 2: DSST</b><br/>N = 2,607"]

        O0 --> O1
        O0 -.- OX1
        O1 --> O2
        O1 -.- OX2
        O2 --> O3
        O2 --> O4
        O2 -.- OX3
    end

    %% TODO: fill in ELISA numbers from the messages in the Elisa chunk above
    subgraph ELISA["<b>Elisa</b>"]
        direction TB
        E0["CHS participants<br/>with Elisa<br/><b>N = xx</b>"]
        EX1["<b>Excluded, n = xx</b><br/>Not in Cognition Study: xx<br/>Prevalent dementia: xx"]
        E1["Dementia-free at baseline<br/><b>N = xx</b>"]
        EX2["<b>Excluded, n = xx</b><br/>Dementia onset &lt; 1 year"]
        E2["<b>Aim 1: Incident dementia</b><br/>N = xx"]
        EX3["<b>Excluded</b><br/>No Y5 score or follow-up<br/>3MSE: xx · DSST: xx"]
        E3["<b>Aim 2: 3MSE</b><br/>N = xx"]
        E4["<b>Aim 2: DSST</b><br/>N = xx"]

        E0 --> E1
        E0 -.- EX1
        E1 --> E2
        E1 -.- EX2
        E2 --> E3
        E2 --> E4
        E2 -.- EX3
    end

    classDef main  fill:#EEF3FA,stroke:#4C78A8,stroke-width:1.5px,color:#1a1a1a;
    classDef aim   fill:#4C78A8,stroke:#35557A,stroke-width:1.5px,color:#ffffff;
    classDef excl  fill:#ffffff,stroke:#9AA3AE,stroke-dasharray:4 3,color:#444444;
    classDef frame fill:#FAFBFC,stroke:#D5DAE1,stroke-width:1px,color:#1a1a1a;

    class S0,S1,O0,O1,E0,E1 main;
    class S2,S3,S4,O2,O3,O4,E2,E3,E4 aim;
    class SX1,SX2,SX3,OX1,OX2,OX3,EX1,EX2,EX3 excl;
    class SOMA,OLINK,ELISA frame;
```

## New Venn

### For aim 1

```{r}
#| fig-width: 6
#| fig-height: 6


ggVennDiagram(list(Olink = idnos_olink_aim1, SomaScan = idnos_soma_aim1, Elisa = idnos_elisa_aim1),
              label_alpha = 0, edge_size = 0.5,
              set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none")
```

### For aim 2

```{r}
#| fig-width: 15
#| fig-height: 5

library(patchwork)

olink <- ggVennDiagram(
  list(
       '3MSE' = idnos_olink_aim2_3mse, 
       'DSST' = idnos_olink_aim2_dsst),
              label_alpha = 0, edge_size = 0.5,
              set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  labs(title = "Olink") +
  theme(legend.position = "none")

soma <- ggVennDiagram(
  list(
    '3MSE' = idnos_soma_aim2_3mse, 
    'DSST' = idnos_soma_aim2_dsst),
  label_alpha = 0, edge_size = 0.5,
  set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  labs(title = "SomaScan") +
  theme(legend.position = "none")

elisa <- ggVennDiagram(
  list(
    '3MSE' = idnos_elisa_aim2_3mse, 
    'DSST' = idnos_elisa_aim2_dsst),
  label_alpha = 0, edge_size = 0.5,
  set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  labs(title = "Elisa") +
  theme(legend.position = "none")

olink + soma + elisa
```

## Table 1

```{r}
data_clin_filt <- data_clin %>%
  filter(idno %in% idnos_olink_aim1 | idno %in% idnos_soma_aim1 | idno %in% idnos_elisa_aim1) %>%
  mutate(
    platform = case_when(
      idno %in% idnos_olink_aim1 & idno %in% idnos_soma_aim1 & idno %in% idnos_elisa_aim1 ~ "Olink, SomaScan and Elisa",
      idno %in% idnos_olink_aim1 & idno %in% idnos_soma_aim1 ~ "Olink and SomaScan",
      idno %in% idnos_olink_aim1 & idno %in% idnos_elisa_aim1 ~ "Olink and Elisa",
      idno %in% idnos_soma_aim1 & idno %in% idnos_elisa_aim1 ~ "SomaScan and Elisa",
      idno %in% idnos_olink_aim1 ~ "Olink only",
      idno %in% idnos_soma_aim1 ~ "SomaScan only",
      idno %in% idnos_elisa_aim1 ~ "Elisa only",
    ))


```

```{r}

ids_soma  <- idnos_soma_aim1
ids_olink <- idnos_olink_aim1
ids_elisa <- idnos_elisa_aim1
ids_all <- Reduce(intersect, list(idnos_soma_aim1, idnos_olink_aim1, idnos_elisa_aim1))


yn <- function(x) if (is.numeric(x)) factor(x, levels = 0:1, labels = c("No", "Yes")) else x

tbl_data <- bind_rows(
  data_clin_filt %>% filter(idno %in% ids_soma)  %>% mutate(platform = "SomaScan"),
  data_clin_filt %>% filter(idno %in% ids_olink) %>% mutate(platform = "Olink"),
  data_clin_filt %>% filter(idno %in% ids_elisa) %>% mutate(platform = "Elisa"),
  data_clin_filt %>% filter(idno %in% ids_all)  %>% mutate(platform = "All platforms")
) %>%
  mutate(
    platform = factor(platform, levels = c("SomaScan", "Olink", "Elisa", "All platforms")),
    apoe4    = factor(if_else(grepl("4", apoecvd), "Yes", "No")),   # NA stays NA
    incident_dem = factor(if_else(demclass == "Incident", "Yes", "No")),
    across(c(prevdm_y5, kiddis_y5, htnmed_y5, lipid_y5, estrgn_y5), yn)
  ) %>%
  select(
    platform,
    age_y5, gend01, race01, grade01, clinic, perstat,
    bmi_y5, waist_y5, smoke_y5, alcoh_y5, kcal_y5,
    avesys_y5, hyper_y5, htnmed_y5, prevdm_y5, gluadj_y5,
    ldl_y5, hdl_y5, trig_y5, lipid_y5, crp_y5,
    egfr_y5, kiddis_y5, chdyr5, stkyr5, estrgn_y5, depscr_y5,
    apoe4, incident_dem, yearsfu
  )

skewed <- c("alcoh_y5", "kcal_y5", "trig_y5", "crp_y5", "gluadj_y5", "yearsfu")

table1 <- tbl_data %>%
  tbl_summary(
    by = platform,
    type = list(c(age_y5, avesys_y5, depscr_y5) ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_of(skewed)   ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    value = list(
      c(htnmed_y5, prevdm_y5, lipid_y5, kiddis_y5, estrgn_y5, apoe4, incident_dem) ~ "Yes",
      gend01 ~ "Male"
    ),
    digits = list(all_continuous() ~ 1, all_categorical() ~ c(0, 1)),
    label = list(
      age_y5       ~ "Age, years",
      gend01       ~ "Male sex",
      race01       ~ "Race",
      grade01      ~ "Education",
      clinic       ~ "Field center",
      perstat      ~ "Cohort",
      bmi_y5       ~ "BMI, kg/m²",
      waist_y5     ~ "Waist circumference, cm",
      smoke_y5     ~ "Smoking status",
      alcoh_y5     ~ "Alcohol, drinks/week",
      kcal_y5      ~ "Physical activity, kcal/week",
      avesys_y5    ~ "Systolic BP, mmHg",
      hyper_y5     ~ "Hypertension status",
      htnmed_y5    ~ "Antihypertensive medication",
      prevdm_y5    ~ "Diabetes",
      gluadj_y5    ~ "Glucose, mg/dL",
      ldl_y5       ~ "LDL cholesterol, mg/dL",
      hdl_y5       ~ "HDL cholesterol, mg/dL",
      trig_y5      ~ "Triglycerides, mg/dL",
      lipid_y5     ~ "Lipid-lowering medication",
      crp_y5       ~ "CRP, mg/L",
      egfr_y5      ~ "eGFR, mL/min/1.73 m²",
      kiddis_y5    ~ "Self-reported kidney disease",
      chdyr5       ~ "Prevalent CHD",
      stkyr5       ~ "Prevalent stroke",
      estrgn_y5    ~ "Estrogen use",
      depscr_y5    ~ "CES-D depression score",
      apoe4        ~ "APOE ε4 carrier",
      incident_dem ~ "Incident dementia during follow-up",
      yearsfu      ~ "Follow-up, years"
    ),
    missing = "ifany",
    missing_text = "Missing"
  ) %>%
  modify_header(
    label ~ "**Characteristic**",
    all_stat_cols() ~ "**{level}**  \nN = {style_number(n)}"
  ) %>%
  modify_spanning_header(all_stat_cols() ~ "**Proteomics platform**") %>%
  modify_caption("**Table 1.** Baseline (CHS Year 5) characteristics by proteomics platform") %>%
  modify_footnote(all_stat_cols() ~
                    "Mean (SD), median (IQR) or n (%). Participants measured on more than one platform appear in each of their platform columns.") %>%
  bold_labels()

table1

# Export
table1 %>% as_gt() %>% gtsave(file.path(data_path, 'Analysis', '1_data_exploration', 'table1.docx' ))   # or .html / .png

```

```{r}
write_csv(data_clin_filt, file.path(data_path, 'Data_processed', 'Meineche_10237_recoded_filtered.csv'))
```

# Descriptives

**2. Descriptives**

- Table 1 across tertiles of the primary exposure.

- Klotho against age, sex and kidney function (Klotho is strongly tied to the kidney, so consider adding eGFR as a covariate).

- Replace the placeholder cognition score in the GAM plot with the real year-5 3MSE and DSST.

## Tertiles of primary exposure

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis


yn <- function(x) if (is.numeric(x)) factor(x, levels = 0:1, labels = c("No", "Yes")) else x

values_klot <- full_join(
  data_olink %>% 
    mutate(idno = as.character(idno)) %>% 
    rename_with(~ paste0(.x, "_olink"), -idno),
  data_soma  %>% 
    mutate(idno = as.character(idno)) %>% 
    rename_with(~ paste0(.x, "_soma"), -idno),
  by = "idno"
  ) %>%
  full_join(
    data_elisa %>% 
      mutate(idno = as.character(idno)) %>% 
      select(idno, KL = klot_elisa) %>% 
      rename_with(~ paste0(.x, "_elisa"), -idno),
    by = "idno"
  ) %>%
  filter(idno %in% c(idnos_olink_aim1, idnos_soma_aim1, idnos_elisa_aim1))

klot_vars <- values_klot %>% select(ends_with("_olink"), ends_with("_soma"), ends_with("_elisa")) %>% names()

# ---- Clinical data with derived variables (built once) ----
clin_tbl <- data_clin_filt %>%
  mutate(
    idno = as.character(idno),
    across(c(prevdm_y5, kiddis_y5, htnmed_y5, lipid_y5, estrgn_y5), yn),
    apoe4 = factor(if_else(grepl("4", apoecvd), "Yes", "No")),
    incident_dem = factor(if_else(demclass == "Incident", "Yes", "No")),
    dem_outcome = case_when(
      demclass == "Incident" & demtype == "AD only"  ~ "Incident: AD",
      demclass == "Incident" & demtype == "Mixed"    ~ "Incident: mixed AD/VaD",
      demclass == "Incident" & demtype == "VaD only" ~ "Incident: VaD",
      demclass == "Incident"                         ~ "Incident: other/unknown type",
      demclass == "MCI"                              ~ "MCI (no dementia)",
      demclass == "Normal"                           ~ "Normal"
    ) %>% factor(levels = c("Normal", "MCI (no dementia)", "Incident: AD",
                            "Incident: mixed AD/VaD", "Incident: VaD",
                            "Incident: other/unknown type"))
  )

tbl_vars <- c(
  "age_y5", "gend01", "race01", "grade01", "clinic", "perstat",
  "bmi_y5", "waist_y5", "smoke_y5", "alcoh_y5", "kcal_y5",
  "avesys_y5", "hyper_y5", "htnmed_y5", "prevdm_y5", "gluadj_y5",
  "ldl_y5", "hdl_y5", "trig_y5", "lipid_y5", "crp_y5",
  "egfr_y5", "kiddis_y5", "chdyr5", "stkyr5", "estrgn_y5", "depscr_y5",
  "apoe4",
  "score134_y5", "digcor_y5",
  "incident_dem", "dem_outcome", "yearsfu"
)

skewed <- c("alcoh_y5", "kcal_y5", "trig_y5", "crp_y5", "gluadj_y5", "yearsfu")


# Loop for each measure
for (klot_var in klot_vars) {

  tbl <- values_klot %>%
    select(idno, all_of(klot_var)) %>%
    filter(!is.na(.data[[klot_var]])) %>%
    left_join(clin_tbl, by = "idno") %>%
    mutate(tertile = ntile(.data[[klot_var]], 3) %>%
             factor(labels = c("Low", "Mid", "High"))) %>%
    select(tertile, all_of(tbl_vars)) %>%
    tbl_summary(
      by = tertile,
      type = list(c(age_y5, avesys_y5, depscr_y5, score134_y5, digcor_y5) ~ "continuous"),
      statistic = list(
        all_continuous()  ~ "{mean} ({sd})",
        all_of(skewed)    ~ "{median} ({p25}, {p75})",
        all_categorical() ~ "{n} ({p}%)"
      ),
      value = list(
        c(htnmed_y5, prevdm_y5, lipid_y5, kiddis_y5, estrgn_y5, apoe4, incident_dem) ~ "Yes",
        gend01 ~ "Male"
      ),
      digits = list(all_continuous() ~ 1, all_categorical() ~ c(0, 1)),
      label = list(
        age_y5       ~ "Age, years",
        gend01       ~ "Male sex",
        race01       ~ "Race",
        grade01      ~ "Education",
        clinic       ~ "Field center",
        perstat      ~ "Cohort",
        bmi_y5       ~ "BMI, kg/m²",
        waist_y5     ~ "Waist circumference, cm",
        smoke_y5     ~ "Smoking status",
        alcoh_y5     ~ "Alcohol, drinks/week",
        kcal_y5      ~ "Physical activity, kcal/week",
        avesys_y5    ~ "Systolic BP, mmHg",
        hyper_y5     ~ "Hypertension status",
        htnmed_y5    ~ "Antihypertensive medication",
        prevdm_y5    ~ "Diabetes",
        gluadj_y5    ~ "Glucose, mg/dL",
        ldl_y5       ~ "LDL cholesterol, mg/dL",
        hdl_y5       ~ "HDL cholesterol, mg/dL",
        trig_y5      ~ "Triglycerides, mg/dL",
        lipid_y5     ~ "Lipid-lowering medication",
        crp_y5       ~ "CRP, mg/L",
        egfr_y5      ~ "eGFR, mL/min/1.73 m²",
        kiddis_y5    ~ "Self-reported kidney disease",
        chdyr5       ~ "Prevalent CHD",
        stkyr5       ~ "Prevalent stroke",
        estrgn_y5    ~ "Estrogen use",
        depscr_y5    ~ "CES-D depression score",
        apoe4        ~ "APOE ε4 carrier",
        score134_y5  ~ "3MSE score, Year 5",
        digcor_y5    ~ "DSST score, Year 5",
        incident_dem ~ "Incident dementia during follow-up",
        dem_outcome  ~ "Dementia status / subtype",
        yearsfu      ~ "Follow-up, years"
      ),
      missing = "ifany",
      missing_text = "Missing"
    ) %>%
    add_p(test = list(all_categorical() ~ "chisq.test")) %>%
    modify_header(
      label ~ "**Characteristic**",
      all_stat_cols() ~ "**{level}**  \nN = {style_number(n)}"
    ) %>%
    modify_spanning_header(all_stat_cols() ~ paste0("**Tertiles of ", klot_var, "**")) %>%
    bold_labels() %>%
    bold_p()
  
  cat("\n\n### ", klot_var, "\n\n", sep = "")
  
  # Save
  tbl %>% as_gt() %>% 
    gtsave(
      file.path(
        data_path, 'Analysis', '1_data_exploration', 'klot_tertiles', 
        glue('table1_tertile_{klot_var}.docx') 
        )) 
  
  # Print in norebook
  print(tbl)
  
}

```

## Klotho against age, sex and kidney function

### Unadjusted

```{r}
#| figure-width: 7
#| figure-height: 4

tbl <- left_join(clin_tbl, values_klot, by = "idno") %>%
  select(idno, all_of(klot_vars), age_y5, gend01, egfr_y5, cystatc_y5, race01, clinic)

cormat <- tbl %>%
  pivot_longer(all_of(klot_vars), names_to = "protein", values_to = "value") %>%
  group_by(protein) %>%
  summarise(
    rho_age    = cor(value, age_y5,     method = "spearman", use = "complete.obs"),
    rho_egfr   = cor(value, egfr_y5,    method = "spearman", use = "complete.obs"),
    rho_cystc  = cor(value, cystatc_y5, method = "spearman", use = "complete.obs"),
    # sex: difference in medians (Male - Female) + Wilcoxon p
    diff_sex   = median(value[gend01 == "Male"], na.rm = TRUE) -
      median(value[gend01 == "Female"], na.rm = TRUE),
    p_sex      = wilcox.test(value ~ gend01)$p.value
  )

hm <- cormat %>%
  pivot_longer(cols = -c(protein, p_sex), 
              values_to = 'rho',
              names_to = 'fac') %>%
  ggplot(aes(x=protein, y=fac, fill=rho)) + 
  geom_tile() + 
  scale_fill_gradient2(low = "steelblue", high = "#990033",mid = "white", limits = c(-1, 1)) +
  geom_text(inherit.aes = T, aes(label = round(rho, 2))) + 
  labs(y = 'Factor', x = 'Protein', fill = '') + 
  theme_minimal()

hm

ggsave(file.path(data_path, 'Analysis', '1_data_exploration', 'cor_hm_clin.png'),
       hm, height = 4, width = 7)  

```

```{r}
#| figure-width: 8
#| figure-height: 8

corplots <- tbl %>%
  select(all_of(klot_vars), age_y5, egfr_y5, cystatc_y5) %>%
  pivot_longer(all_of(klot_vars), names_to = "protein", values_to = "value") %>%
  pivot_longer(c(age_y5, egfr_y5, cystatc_y5), names_to = "fac", values_to = "level") %>%
  ggplot(aes(x = level, y = value)) +
  geom_point(alpha = 0.1, size = 0.4, colour = "grey50") +
  geom_smooth(method = "lm", colour = "#4C78A8", fill = "#4C78A8", alpha = 0.2) +
  stat_cor(method = "spearman", cor.coef.name = "rho",
           label.x.npc = "left", label.y.npc = "top",
           size = 3, p.accuracy = 0.001, r.accuracy = 0.01) +
  facet_grid(protein ~ fac, scales = "free") +
  labs(x = NULL, y = "Protein level") +
  theme_minimal()

corplots

ggsave(file.path(data_path, 'Analysis', '1_data_exploration', 'corplots_num_klot.png'),
       corplots, height = 8, width = 8)  

```

```{r}
#| figure-width: 10
#| figure-height: 12

boxplots <- tbl %>%
  select(c(all_of(klot_vars), race01, clinic, gend01)) %>%
  pivot_longer(all_of(klot_vars), 
               names_to = "protein", 
               values_to = "value") %>%
  pivot_longer(c(race01, clinic, gend01),
                names_to  = "fac",
                values_to = "level",
                values_transform = list(level = as.character)) %>%
  ggplot(aes(x=level, y=value, fill = level)) + 
  geom_boxplot(outlier.size = 0.4) + 
  facet_grid(protein ~ fac , scales = "free") + 
  scale_fill_brewer(palette = "Set2") +
  labs(y='Intensity', x='s') + 
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")


boxplots +
  geom_pwc(
    method = "wilcox_test",           # pairwise Mann–Whitney (skewed data)
    label = "p.adj.signif",           # * / ** / *** after adjustment
    p.adjust.method = "holm",
    hide.ns = TRUE,                   # only draw brackets that are significant
    tip.length = 0.01,
    step.increase = 0.08,
    size = 0.3, label.size = 3
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))   # room for brackets

boxplots

ggsave(file.path(data_path, 'Analysis', '1_data_exploration', 'boxplot_clinic_race_klot.png'),
       boxplots, height = 12, width = 10)  
```

### Mutually adjusted

```{r}
library(broom)

assoc_res <- map_dfr(klot_vars, function(v) {
  lm(scale(tbl[[v]]) ~ scale(age_y5) + gend01 + scale(egfr_y5) + race01 + clinic,
     data = tbl) %>%
    tidy(conf.int = TRUE) %>%
    filter(term %in% c("scale(age_y5)", "gend01Male", "scale(egfr_y5)")) %>%
    mutate(protein = v)
})

assoc_res


write_xlsx(assoc_res, file.path(data_path, 'Analysis', '1_data_exploration', 'lm_age_eGFR_sex.xlsx'))

# Forest plot

assoc_res %>%
  mutate(term = recode(term,
                       "scale(age_y5)"  = "Age (per SD)",
                       "gend01Male"     = "Male vs female",
                       "scale(egfr_y5)" = "eGFR (per SD)")) %>%
  ggplot(aes(x = estimate, y = protein, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_pointrange(colour = "#4C78A8") +
  facet_wrap(~ term) +
  labs(x = "Difference in protein (SD units), mutually adjusted", y = NULL) +
  theme_minimal()

ggsave(file.path(data_path, 'Analysis', '1_data_exploration', 'lm_age_eGFR_sex.png'))
```

## GAM of cognition

Replace the placeholder cognition score in the GAM plot with the real year-5 3MSE and DSST

```{r}
#| fig-width: 10
#| fig-height: 12
#| message: false
#| warning: false


gam_plot <- left_join(clin_tbl, values_klot, by = "idno") %>%
  select(idno, all_of(klot_vars), score134_y5, digcor_y5) %>%
  pivot_longer(all_of(klot_vars), names_to = "protein", values_to = "value") %>%
  pivot_longer(c(score134_y5, digcor_y5), names_to = "test", values_to = "score") %>%
  mutate(test = recode(test, score134_y5 = "3MSE, Year 5", digcor_y5 = "DSST, Year 5")) %>%
  filter(!is.na(value), !is.na(score)) %>%
  ggplot(aes(x = value, y = score)) +
  geom_point(alpha = 0.1, size = 0.4, colour = "grey50") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              colour = "#4C78A8", fill = "#4C78A8", alpha = 0.2) +
  facet_grid(protein ~ test, scales = "free") +
  labs(x = "Protein level", y = "Cognitive score") +
  theme_minimal()

gam_plot
```

```{r}

data_soma_win <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'soma_y5_klotho.xlsx'
  ), 
  sheet = 'values_win99'
) %>% mutate(idno=as.character(idno))

data_olink_win <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'olink_y5_klotho.xlsx'
  ), 
  sheet = 'values_win99'
)

data_elisa_win <- read_xlsx(
  file.path(
    data_path, 'Data_processed', 'elisa_klotho.xlsx'
  ), 
  sheet = 'values_win99'
)

values_klot_win <- full_join(data_olink_win %>% 
                               mutate(idno = as.character(idno)) %>% 
                               rename_with(~ paste0(.x, "_olink"), -idno),
                             data_soma_win  %>% 
                               mutate(idno = as.character(idno)) %>% 
                               rename_with(~ paste0(.x, "_soma"), -idno),
                             by = "idno") %>%
  full_join(data_elisa_win %>% 
              mutate(idno = as.character(idno)) %>% 
              select(idno, KL = klot_elisa) %>% 
              rename_with(~ paste0(.x, "_elisa"), -idno),
            by = "idno")

gam_plot <- left_join(clin_tbl, values_klot_win, by = "idno") %>%
  select(idno, all_of(klot_vars), score134_y5, digcor_y5) %>%
  pivot_longer(all_of(klot_vars), names_to = "protein", values_to = "value") %>%
  pivot_longer(c(score134_y5, digcor_y5), names_to = "test", values_to = "score") %>%
  mutate(test = recode(test, score134_y5 = "3MSE, Year 5", digcor_y5 = "DSST, Year 5")) %>%
  filter(!is.na(value), !is.na(score)) %>%
  ggplot(aes(x = value, y = score)) +
  geom_point(alpha = 0.1, size = 0.4, colour = "grey50") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              colour = "#4C78A8", fill = "#4C78A8", alpha = 0.2)+
  stat_cor(method = "spearman", cor.coef.name = "rho",
           label.x.npc = "left", label.y.npc = "bottom",
           size = 3, r.accuracy = 0.01, p.accuracy = 0.001)+
  facet_grid(test~protein, scales = "free") +
  labs(x = "Protein level", y = "Cognitive score") +
  theme_minimal()

gam_plot

ggsave(file.path(data_path, 'Analysis', '1_data_exploration', 'gam_prot_cog.png'),
       gam_plot, height = 5, width = 9)
```
