---
title: "Klotho, dementia, cognition and genetics in CHS"
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: katex
    grid:
      body-width: 1200px
---

# Introduction

This notebook analyses plasma α-Klotho measured at CHS Year 5. The measurement method (e.g. ELISA, SOMA, Olink) is set by `measure_method` in Setup, and all output goes to a folder named after it.

- **Descriptives:** baseline characteristics (Table 1), characteristics by klotho tertile (Table 2), and klotho levels by APOE genotype and by prevalent dementia.
- **Aim 1:** klotho and incident dementia (Cox models).
- **Aim 2:** klotho and cognitive trajectories, 3MSE and DSST (linear mixed models).
- **Aim 3:** genetic variants in *KL* and klotho levels, including the KL-VS haplotype and its association with dementia.

All output is written to `data_path/Analysis/measurement method/`:

| Folder | Content |
|:----------------------|:------------------------------------------------|
| `01_descriptives/` | Tables 1–2, boxplots and APOE comparisons |
| `02_cox_dementia/` | Cox models, one folder per `<outcome>/<covariate set>/` |
| `03_lmm_cognition/` | Mixed models, one folder per `<test>/<covariate set>/` |
| `04_genetics/` | SNP analyses, with KL-VS results in `klvs/` |

# Setup

## Packages and paths

`paths.R` defines `data_path`.

```{r}
#| message: false

rm(list = ls())

library(tidyverse)
library(readxl)
library(writexl)
library(glue)
library(gt)
library(gtsummary)
library(ggpubr)
library(ggVennDiagram)
library(patchwork)
library(survival)
library(ggsurvfit)
library(broom)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggeffects)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)


source("paths.R")
source("forest_lmm.R")   # plot_forest_lmm(), used to combine the mixed-model results

measure_method = 'ELISA'

dir_out   <- file.path(data_path, "Analysis", measure_method)
dir_desc  <- file.path(dir_out, "01_descriptives")
dir_cox   <- file.path(dir_out, "02_cox_dementia")
dir_lmm   <- file.path(dir_out, "03_lmm_cognition")
dir_gen   <- file.path(dir_out, "04_genetics")
dir_klvs  <- file.path(dir_gen, "klvs")

walk(c(dir_desc, dir_cox, dir_lmm, dir_gen, dir_klvs),
     dir.create, recursive = TRUE, showWarnings = FALSE)
```

**Measurement method:** `r measure_method`

## Covariate sets

Each model adds covariates to the previous one.

- `cov_m0`: crude (exposure only)
- `cov_m1`: demographics, kidney function, clinic, education and APOE ε4
- `cov_m2`: Model 1 + cardiovascular risk factors and lifestyle
- `cov_m3`: Model 2 + CRP, depressive symptoms, self-rated health and estrogen use

```{r}
cov_m0 <- c()
cov_m1 <- c("age_y5", "gend01", "egfr_y5", "race01", "clinic", "grade01", "apoe4")
cov_m2 <- c(cov_m1, "bmi_y5", "smoke_y5", "alcoh_y5", "kcal_y5",
            "avesys_y5", "htnmed_y5", "prevdm_y5", "ldl_y5", "hdl_y5",
            "lipid_y5", "chdyr5", "stkyr5")
cov_m3 <- c(cov_m2, "crp_y5", "depscr_y5", "hlth1_y5", "estrgn_y5")

# NOTE: cov_m05 is used in Aim 3 (KL-VS) but is not defined here yet.
# cov_m05 <- c(...)
```

## Labels

Readable names for model terms (forest plots) and for covariates (adjustment descriptions).

```{r}
term_labels <- c(
  klot_value                     = "Klotho (per SD log2)",
  `KL_VSHet+`                    = "KL-VS heterozygous (vs non-heterozygous)",
  age_y5                         = "Age (per year)",
  gend01Male                     = "Male sex",
  egfr_y5                        = "eGFR (per mL/min/1.73 m²)",
  race01Black                    = "Race: Black (vs White)",
  race01Other                    = "Race: Other (vs White)",
  clinicDavis                    = "Center: Davis",
  clinicHagerstown               = "Center: Hagerstown",
  clinicPittsburgh               = "Center: Pittsburgh",
  `grade01Less than high school` = "Less than high school",
  apoe4Yes                       = "APOE ε4 carrier",
  bmi_y5                         = "BMI (per kg/m²)",
  alcoh_y5                       = "Alcohol (per drink/week)",
  kcal_y5                        = "Physical activity (per kcal/week)",
  avesys_y5                      = "Systolic BP (per mmHg)",
  htnmed_y5                      = "Antihypertensive medication",
  prevdm_y5                      = "Diabetes",
  ldl_y5                         = "LDL cholesterol (per mg/dL)",
  hdl_y5                         = "HDL cholesterol (per mg/dL)",
  lipid_y5Yes                    = "Lipid-lowering medication",
  chdyr5Yes                      = "Prevalent CHD",
  stkyr5Yes                      = "Prevalent stroke",
  crp_y5Yes                      = "CRP (per mg/L)",
  depscr_y5                      = "CES-D score (per point)",
  estrgn_y5Yes                   = "Estrogen use",
  `smoke_y5Former smoker`        = "Smoking: former",
  `smoke_y5Never smoked`         = "Smoking: never"
)

cov_labels <- c(
  age_y5 = "age", gend01 = "sex", race01 = "race", egfr_y5 = "eGFR",
  clinic = "clinic", grade01 = "education", apoe4 = "APOE ε4",
  bmi_y5 = "BMI", smoke_y5 = "smoking", alcoh_y5 = "alcohol", kcal_y5 = "physical activity",
  avesys_y5 = "systolic BP", htnmed_y5 = "antihypertensives", prevdm_y5 = "diabetes",
  ldl_y5 = "LDL", hdl_y5 = "HDL", lipid_y5 = "lipid-lowering drugs",
  chdyr5 = "prevalent CHD", stkyr5 = "prevalent stroke",
  crp_y5 = "CRP", depscr_y5 = "CES-D", hlth1_y5 = "self-rated health", estrgn_y5 = "estrogen use"
)
nice_cov <- function(x) coalesce(unname(cov_labels[x]), x)   # unknown names stay as they are
```

## Functions

**Formatting and forest plots.** `format_forest()` adds labels, the estimate with 95% CI, and formatted p-values to a model results table. `plot_forest_hr()` draws hazard ratios on a log scale (Cox models); `plot_forest_coef()` draws linear estimates (mixed models).

```{r}
fmt_p <- function(p) if_else(p < 0.001, "< 0.001", sprintf("= %.3f", p))

format_forest <- function(results, est = "HR", drop_terms = character(0)) {
  results %>%
    filter(!term %in% drop_terms) %>%
    mutate(
      label  = coalesce(term_labels[term], term),           # unknown terms keep their raw name
      label  = factor(label, levels = rev(unique(label))),  # keep model order, top to bottom
      est_ci = sprintf("%.2f (%.2f–%.2f)", .data[[est]], lo, hi),
      p_lab  = if_else(p < 0.001, "<0.001", sprintf("%.3f", p))
    )
}

plot_forest_hr <- function(data_plot) {
  ggplot(data_plot, aes(x = HR, y = label)) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2, colour = "#2171B5") +
    geom_point(size = 2.5, colour = "#2171B5") +
    geom_text(aes(x = max(hi) * 1.3, label = paste0(est_ci, "   p = ", p_lab)),
              hjust = 0, size = 3.2) +
    scale_x_log10(expand = expansion(mult = c(0.05, 0.9))) +
    labs(x = "Hazard ratio (95% CI), log scale", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}

plot_forest_coef <- function(data_plot, title) {
  ggplot(data_plot, aes(x = estimate, y = label)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2, colour = "#2171B5") +
    geom_point(size = 2.5, colour = "#2171B5") +
    geom_text(aes(x = Inf, label = paste0(est_ci, "   p = ", p_lab)),
              hjust = 0, size = 3.2) +
    coord_cartesian(clip = "off") +
    labs(x = "Estimate (95% CI)", y = NULL, title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.margin = margin(5.5, 170, 5.5, 5.5))   # right margin for the text
}
```

**Cox model.** `run_cox()` fits

$$\text{Surv}(\text{yearsfu}, \text{outcome}) \sim \text{exposure} + \text{covariates}$$

with robust standard errors, and saves to `<out_dir>/<outcome>/<model_name>/`:

- the model summary and a table of hazard ratios, model fit (N, events, C-index, AIC) and the proportional hazards test (`Cox_results.xlsx`),
- the Schoenfeld residual plot for the exposure (flat line = constant HR over time),
- for a numeric exposure, a penalised spline model where the `nonlin` row tests departure from linearity.

`save_forest_cox()` draws the forest plot of all terms and saves it in the same folder.

```{r}
run_cox <- function(
    outcome,
    covar,
    data                 = data_combined_aim1,
    variable_of_interest = "klot_value",
    model_name           = deparse(substitute(covar)),
    out_dir              = dir_cox
) {
  
  save_dir <- file.path(out_dir, outcome, model_name)
  dir.create(save_dir, recursive = TRUE, showWarnings = FALSE)
  message(glue("\n========== {outcome} – {model_name} =========="))
  message(glue("Saving to {save_dir}"))
  
  is_num <- is.numeric(data[[variable_of_interest]])
  
  # ---- main model ----
  modelForm <- as.formula(
    paste0("Surv(yearsfu, ", outcome, ") ~ ",
           paste(c(variable_of_interest, covar), collapse = " + "))
  )
  message(glue("Formula: {deparse1(modelForm)}"))
  
  fit <- coxph(modelForm, data = data, id = idno, robust = TRUE)
  message(glue("N = {fit$n}, events = {fit$nevent}"))
  print(fit)
  capture.output(summary(fit), file = file.path(save_dir, "cox_summary.txt"))
  
  results <- tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    select(term, HR = estimate, lo = conf.low, hi = conf.high, p = p.value)
  print(results)
  
  model_info <- glance(fit) %>%
    select(n, nevent, concordance, std.error.concordance, AIC)
  print(model_info)
  
  # ---- proportional hazards ----
  zph <- cox.zph(fit)
  print(zph)                                       # p < 0.05 = PH violated
  zph_tab <- as.data.frame(zph$table) %>% rownames_to_column("term")
  
  draw_zph <- function() {
    plot(zph, var = variable_of_interest,
         main = glue("Schoenfeld residuals – {model_name}"))
    if (is_num) abline(h = coef(fit)[variable_of_interest], lty = 2, col = "red")
  }
  draw_zph()
  png(file.path(save_dir, "ph_schoenfeld.png"), width = 7, height = 5, units = "in", res = 300)
  draw_zph()
  dev.off()
  
  # ---- linearity (numeric exposure only) ----
  fit_spl <- NULL
  if (is_num) {
    fit_spl <- coxph(
      as.formula(paste0("Surv(yearsfu, ", outcome, ") ~ ",
                        paste(c(glue("pspline({variable_of_interest}, df = 3)"), covar),
                              collapse = " + "))),
      data = data
    )
    print(fit_spl)                                 # "nonlin" row tests departure from linearity
    capture.output(print(fit_spl), file = file.path(save_dir, "spline_summary.txt"))
    
    draw_spl <- function() {
      termplot(fit_spl, term = 1, se = TRUE, rug = TRUE,
               ylabs = "log HR", xlabs = variable_of_interest,
               main = glue("Spline – {model_name}"))
    }
    draw_spl()
    png(file.path(save_dir, "spline_termplot.png"), width = 7, height = 5, units = "in", res = 300)
    draw_spl()
    dev.off()
  }
  
  # ---- save tables ----
  write_xlsx(
    x    = list(HR = results, model_info = model_info, PH_test = zph_tab),
    path = file.path(save_dir, "Cox_results.xlsx")
  )
  
  invisible(list(fit = fit, results = results, model_info = model_info,
                 zph = zph, fit_spl = fit_spl, save_dir = save_dir))
}

save_forest_cox <- function(res, drop_terms = character(0), ...) {
  p <- res$results %>%
    format_forest("HR", drop_terms) %>%
    plot_forest_hr()
  ggsave(file.path(res$save_dir, "forest.png"), p, bg = "white", ...)
  p
}
```

**Linear mixed model.** `run_lmm()` fits

$$\text{score} \sim \text{klotho} \times \text{time} + \text{covariates} + (1 + \text{time} \mid \text{idno})$$

with `time` = years since Year 5 and a random intercept and slope per participant. The two key terms are `klot_value` (difference in score at Year 5) and `klot_value:time` (difference in yearly change). It also plots predicted trajectories at the 10th, 50th and 90th percentiles of klotho and residual diagnostics, and saves everything to `03_lmm_cognition/<outcome>/<model_name>/`.

`run_lmm_models()` runs the full set of models for one cognitive test (Models 0–3 and Model 1 stratified by APOE ε4), and saves a forest plot of all terms for each.

```{r}
run_lmm <- function(
    outcome       = "score134",
    covar,
    ids,
    data          = data_clin_long,
    baseline_year = 5,
    model_name    = deparse(substitute(covar)),
    out_dir       = dir_lmm
) {
  
  save_dir <- file.path(out_dir, outcome, model_name)
  dir.create(save_dir, recursive = TRUE, showWarnings = FALSE)
  message(glue("\n========== {outcome} – {model_name} =========="))
  message(glue("Saving to {save_dir}"))
  
  # ---- data: from baseline onwards, time since Year 5 ----
  data <- data %>%
    filter(idno %in% ids, year >= baseline_year) %>%
    mutate(time = year - baseline_year) %>%
    drop_na(all_of(c(outcome, "klot_value", covar)))
  
  n_ids <- n_distinct(data$idno)
  n_obs <- nrow(data)
  message(glue("N persons = {n_ids}, N observations = {n_obs} ({round(n_obs / n_ids, 1)} per person)"))
  
  # ---- model ----
  modelForm <- as.formula(
    paste(outcome, "~",
          paste(c("klot_value * time", covar, "(1 + time | idno)"), collapse = " + "))
  )
  message(glue("Formula: {deparse1(modelForm)}"))
  
  fit <- lmerTest::lmer(modelForm, data = data)
  print(summary(fit))
  capture.output(summary(fit), file = file.path(save_dir, "lmm_summary.txt"))
  
  results <- tidy(fit, effects = "fixed", conf.int = TRUE) %>%
    select(term, estimate, lo = conf.low, hi = conf.high, p = p.value)
  print(results)
  
  random_eff <- as.data.frame(VarCorr(fit))
  print(random_eff)
  
  model_info <- tibble(
    n_ids, n_obs, obs_per_id = n_obs / n_ids,
    AIC = AIC(fit), BIC = BIC(fit), singular = isSingular(fit)
  )
  print(model_info)
  if (isSingular(fit)) message("WARNING: singular fit – random slope variance may be ~0")
  
  # ---- key result ----
  key <- results %>% filter(term %in% c("klot_value", "time", "klot_value:time"))
  message("Key terms (klot_value = level at Y5, klot_value:time = difference in yearly change):")
  print(key)
  
  # ---- predicted trajectories at 10th / 50th / 90th percentile of klotho ----
  q    <- round(quantile(data$klot_value, c(0.10, 0.50, 0.90)))
  pred <- ggpredict(fit, terms = c("time", glue("klot_value [{paste(q, collapse = ',')}]")))
  
  p_traj <- as_tibble(pred) %>%
    mutate(group = factor(group, labels = c("10th pct", "Median", "90th pct"))) %>%
    ggplot(aes(x = x + baseline_year, y = predicted, colour = group, fill = group)) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 1) +
    scale_colour_manual(values = c("#C6DBEF", "#6BAED6", "#2171B5"), name = "Klotho") +
    scale_fill_manual(values   = c("#C6DBEF", "#6BAED6", "#2171B5"), name = "Klotho") +
    labs(title    = glue("Predicted {outcome} by klotho – {model_name}"),
         subtitle = glue("Klotho × time p = {signif(key$p[key$term == 'klot_value:time'], 2)}"),
         x = "CHS study year", y = glue("Predicted {outcome}")) +
    theme_minimal(base_size = 12)
  print(p_traj)
  ggsave(file.path(save_dir, "trajectories.png"), p_traj, width = 7, height = 5, dpi = 300, bg = "white")
  
  # ---- diagnostics ----
  diag_df <- tibble(fitted = fitted(fit), resid = resid(fit))
  p_diag <-
    (ggplot(diag_df, aes(fitted, resid)) +
       geom_point(alpha = 0.15, size = 0.6) +
       geom_hline(yintercept = 0, linetype = "dashed") +
       geom_smooth(method = "loess", formula = y ~ x, colour = "#2171B5") +
       labs(title = "Residuals vs fitted", x = "Fitted", y = "Residual") +
       theme_minimal()) +
    (ggplot(diag_df, aes(sample = resid)) +
       stat_qq(alpha = 0.15, size = 0.6) + stat_qq_line(colour = "#2171B5") +
       labs(title = "Normal QQ plot of residuals") +
       theme_minimal())
  print(p_diag)
  ggsave(file.path(save_dir, "diagnostics.png"), p_diag, width = 10, height = 4.5, dpi = 300, bg = "white")
  
  # ---- save tables ----
  write_xlsx(
    x = list(fixed = results, random = random_eff, model_info = model_info),
    path = file.path(save_dir, "LMM_results.xlsx")
  )
  
  invisible(list(fit = fit, results = results, key = key,
                 random = random_eff, model_info = model_info, data = data,
                 save_dir = save_dir))
}

save_forest_lmm <- function(res, title) {
  p <- res$results %>%
    format_forest("estimate", drop_terms = "(Intercept)") %>%
    plot_forest_coef(title)
  ggsave(file.path(res$save_dir, "forest.png"), p, bg = "white")
  p
}

run_lmm_models <- function(outcome, ids, test_label) {
  
  cov_m1_noapoe <- setdiff(cov_m1, "apoe4")
  
  res <- list(
    m0 = run_lmm(outcome, cov_m0, ids, model_name = "cov_m0"),
    m1 = run_lmm(outcome, cov_m1, ids, model_name = "cov_m1"),
    m1_apoe4_carriers    = run_lmm(outcome, cov_m1_noapoe, ids,
                                   data = data_clin_long %>% filter(apoe4 == "Yes"),
                                   model_name = "cov_m1_apoe4_carriers"),
    m1_apoe4_noncarriers = run_lmm(outcome, cov_m1_noapoe, ids,
                                   data = data_clin_long %>% filter(apoe4 == "No"),
                                   model_name = "cov_m1_apoe4_noncarriers"),
    m2 = run_lmm(outcome, cov_m2, ids, model_name = "cov_m2"),
    m3 = run_lmm(outcome, cov_m3, ids, model_name = "cov_m3")
  )
  
  titles <- c(m0 = "crude", m1 = "Model 1",
              m1_apoe4_carriers = "Model 1, APOE ε4 carriers",
              m1_apoe4_noncarriers = "Model 1, APOE ε4 non-carriers",
              m2 = "Model 2", m3 = "Model 3")
  
  iwalk(res, \(r, nm) print(save_forest_lmm(r, paste0(test_label, ", ", titles[[nm]]))))
  
  invisible(res)
}
```

# Data

## Clinical data

Education is collapsed into two levels, race and self-reported kidney disease are made factors (White and No as reference), and `idno` is stored as character for joining.

```{r}
data_clin <- read_csv(
  file.path(data_path, "Data_processed", "Meineche_10237_recoded.csv")
) %>%
  mutate(
    grade01 = case_when(
      grade01 == "High school" | grade01 == "College or above" ~ "Highschool or above",
      TRUE ~ grade01
    ),
    race01    = factor(race01, levels = c("White", "Black", "Other")),
    kiddis_y5 = factor(kiddis_y5, levels = c(0, 1), labels = c("No", "Yes")),
    idno      = as.character(idno)
  )
```

## Klotho data

Scaled klotho values (sheet `values_scaled`). `idnos_klot` holds everyone with a klotho measurement, the starting population for all analyses.

```{r}
if (measure_method=='ELISA') {
  data_klot <- read_xlsx(
    file.path(data_path, "Data_processed", "elisa_klotho.xlsx"),
    sheet = "values_scaled"
  ) %>%
    rename(
      klot_value = klot_elisa
    )
  
  idnos_klot <- data_klot %>% pull(idno)
} 

if (measure_method=='SOMA') {
  data_klot <- read_xlsx(
    file.path(data_path, "Data_processed", "soma_y5_klotho.xlsx"),
    sheet = "values_scaled"
  ) %>%
    select(c(idno, KL)) %>%
    rename(
      klot_value = KL
    )
  
  idnos_klot <- data_klot %>% pull(idno)
} 

if (!measure_method %in% c('ELISA', 'SOMA')) {
  stop(glue("No klotho data defined for measure_method = '{measure_method}'."))
}

```

# Participants

Each analysis population is stored as a vector of IDs (`idnos_*`).

## Aim 1: Dementia

Participants with klotho, excluding those not in the cognition study (no dementia information) and those with prevalent dementia. For a sensitivity analysis, those with dementia within the first year of follow-up are also excluded.

```{r}

data_aim1 <- data_clin %>% filter(idno %in% idnos_klot)
n_start   <- nrow(data_aim1)
lab_start <- glue("Start: N = {n_start}")
message(lab_start)

data_aim1 <- data_aim1 %>% filter(!is.na(demclass))
n_cog     <- nrow(data_aim1)
lab_cog   <- glue("{n_start - n_cog} not in cognition study (no dementia information).")
message(lab_cog)

data_aim1 <- data_aim1 %>% filter(demclass != "Prevalent")
n_prev    <- nrow(data_aim1)
idnos_klot_aim1 <- data_aim1 %>% pull(idno)
lab_prev  <- glue("{n_cog - n_prev} removed due to prevalent dementia.")
message(glue("{lab_prev} N = {n_prev}."))

data_aim1_oneyear <- data_aim1 %>% filter(yearsfu > 1)
idnos_klot_aim1_oneyear <- data_aim1_oneyear %>% pull(idno)
n_one     <- nrow(data_aim1_oneyear)
lab_one   <- glue("Sensitivity: {n_prev - n_one} removed due to dementia within the first year.")
message(glue("{lab_one} N = {n_one}."))


# Flowchart built from the same labels and counts
flow_aim1 <- grViz(glue('
digraph flow {{
  graph [rankdir = TB, splines = ortho, nodesep = 0.6]
  node  [shape = box, fontname = Helvetica, fontsize = 11]

  start [label = "{lab_start}"]
  cog   [label = "N = {n_cog}"]
  prev  [label = "N = {n_prev}", style = bold]
  one   [label = "N = {n_one}"]

  ex1 [label = "{lab_cog}"]
  ex2 [label = "{lab_prev}"]
  ex3 [label = "{lab_one}"]

  node [shape = point, width = 0, height = 0]
  p1; p2; p3

  start -> p1 [arrowhead = none]; p1 -> cog
  cog   -> p2 [arrowhead = none]; p2 -> prev
  prev  -> p3 [arrowhead = none]; p3 -> one
  p1 -> ex1; p2 -> ex2; p3 -> ex3

  {{ rank = same; p1; ex1 }}
  {{ rank = same; p2; ex2 }}
  {{ rank = same; p3; ex3 }}
}}
'))

flow_aim1

svg <- export_svg(flow_aim1)
rsvg_png(charToRaw(svg), file.path(dir_desc, "flowchart_aim1.png")) 

rm(list = c("data_aim1", "data_aim1_oneyear", "n_start", "n_cog", "n_prev", "n_one",
            ls(pattern = "^lab_")))

```

## Aim 2: Cognition

Participants with klotho, excluding those with prevalent dementia (those without dementia information are kept). For each test, participants need a Year 5 score and at least one follow-up score (Years 6–11).

```{r}
library(dplyr)
library(glue)
library(DiagrammeR)

data_aim2 <- data_clin %>% filter(idno %in% idnos_klot)
n_start   <- nrow(data_aim2)
lab_start <- glue("Start: N = {n_start}")
message(lab_start)

fu_3mse <- paste0("score134_y", 6:11)
fu_dsst <- paste0("digcor_y", 6:11)

aim2_3mse <- data_aim2 %>%
  filter(!is.na(score134_y5)) %>%
  filter(if_any(all_of(fu_3mse), ~ !is.na(.x)))

aim2_dsst <- data_aim2 %>%
  filter(!is.na(digcor_y5)) %>%
  filter(if_any(all_of(fu_dsst), ~ !is.na(.x)))

idnos_klot_aim2_3mse <- aim2_3mse %>% pull(idno)
idnos_klot_aim2_dsst <- aim2_dsst %>% pull(idno)
n_3mse <- nrow(aim2_3mse)
n_dsst <- nrow(aim2_dsst)

lab_3mse <- glue("3MSE: {n_start - n_3mse} removed (no Y5 score or no follow-up).")
lab_dsst <- glue("DSST: {n_start - n_dsst} removed (no Y5 score or no follow-up).")
message(glue("{lab_3mse} N = {n_3mse}"))
message(glue("{lab_dsst} N = {n_dsst}"))

# Sensitivity

idnos_klot_aim2_3mse_noprevdem <- aim2_3mse %>% 
  filter(is.na(demclass) | demclass != "Prevalent") %>% 
  pull(idno)
n_3mse_sens   <- length(idnos_klot_aim2_3mse_noprevdem)
lab_3mse_sens <- glue("Sensitivity, 3MSE: {n_3mse - n_3mse_sens} removed due to prevalent dementia.")
message(glue("{lab_3mse_sens} N = {n_3mse_sens}"))

idnos_klot_aim2_dsst_noprevdem <- aim2_dsst %>% 
  filter(is.na(demclass) | demclass != "Prevalent") %>% 
  pull(idno)
n_dsst_sens   <- length(idnos_klot_aim2_dsst_noprevdem)
lab_dsst_sens <- glue("Sensitivity, DSST: {n_dsst - n_dsst_sens} removed due to prevalent dementia.")
message(glue("{lab_dsst_sens} N = {n_dsst_sens}"))

# Flowchart

flow_aim2 <- grViz(glue('
digraph flow {{
  graph [rankdir = TB, splines = ortho, nodesep = 0.4, ranksep = 0.35, newrank = true]
  node  [shape = box, fontname = Helvetica, fontsize = 11]

  start [label = "{lab_start}"]
  m1    [label = "N = {n_3mse}", style = bold]
  d1    [label = "N = {n_dsst}", style = bold]
  m2    [label = "N = {n_3mse_sens}"]
  d2    [label = "N = {n_dsst_sens}"]

  exm1  [label = "{lab_3mse}"]
  exd1  [label = "{lab_dsst}"]
  exm2  [label = "{lab_3mse_sens}"]
  exd2  [label = "{lab_dsst_sens}"]

  node [shape = point, width = 0, height = 0]
  pm0; pd0; pm1; pd1; pm2; pd2

  start -> pm0 [arrowhead = none]; start -> pd0 [arrowhead = none]
  pm0 -> pm1 [arrowhead = none];   pd0 -> pd1 [arrowhead = none]
  pm1 -> m1;  pd1 -> d1
  m1 -> pm2 [arrowhead = none]; pm2 -> m2
  d1 -> pd2 [arrowhead = none]; pd2 -> d2

  # 3MSE exclusions to the left, DSST exclusions to the right
  exm1 -> pm1 [dir = back]; pd1 -> exd1
  exm2 -> pm2 [dir = back]; pd2 -> exd2

  {{ rank = same; pm0; pd0; pm0 -> pd0 [style = invis] }}
  {{ rank = same; exm1; pm1; pd1; exd1; exm1 -> pm1 -> pd1 -> exd1 [style = invis] }}
  {{ rank = same; exm2; pm2; pd2; exd2; exm2 -> pm2 -> pd2 -> exd2 [style = invis] }}
  {{ rank = same; m1; d1 }}
  {{ rank = same; m2; d2 }}
}}
'))

flow_aim2

# Save
svg <- DiagrammeRsvg::export_svg(flow_aim2)
rsvg_png(charToRaw(svg), file.path(dir_desc, "flowchart_aim2.png")) 

# Clean up
rm(list = c("data_aim2", "aim2_3mse", "aim2_dsst", "fu_3mse", "fu_dsst", "svg",
            ls(pattern = "^n_"), ls(pattern = "^lab_")))
```

## Overlap between analyses

```{r}
#| fig-width: 6
#| fig-height: 6

ggVennDiagram(list("Aim 1: Dementia" = idnos_klot_aim1,
                   "Aim 2: 3MSE"     = idnos_klot_aim2_3mse,
                   "Aim 2: DSST"     = idnos_klot_aim2_dsst),
              label_alpha = 0, edge_size = 0.5,
              set_size = 4.5, set_color = "grey25") +
  scale_fill_gradient(low = "#F4F6FA", high = "#4C78A8") +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none")
```

# Descriptives

## Table 1: Baseline characteristics

Baseline (Year 5) characteristics by analysis population. Participants appear in every column they belong to; "Total" is everyone in at least one analysis.

```{r}
yn <- function(x) if (is.numeric(x)) factor(x, levels = 0:1, labels = c("No", "Yes")) else x

data_clin_filt <- data_clin %>%
  filter(idno %in% idnos_klot_aim1 |
           idno %in% idnos_klot_aim2_3mse |
           idno %in% idnos_klot_aim2_dsst)

tbl_data <- bind_rows(
  data_clin_filt %>% filter(idno %in% idnos_klot_aim1)      %>% mutate(sample = "Dementia"),
  data_clin_filt %>% filter(idno %in% idnos_klot_aim2_3mse) %>% mutate(sample = "3MSE"),
  data_clin_filt %>% filter(idno %in% idnos_klot_aim2_dsst) %>% mutate(sample = "DSST"),
  data_clin_filt                                              %>% mutate(sample = "Total")
) %>%
  mutate(
    sample = factor(sample, levels = c("Dementia", "3MSE", "DSST", "Total")),
    across(c(prevdm_y5, kiddis_y5, htnmed_y5, lipid_y5, estrgn_y5), yn)
  ) %>%
  select(
    sample,
    age_y5, gend01, race01, grade01,
    bmi_y5, waist_y5, smoke_y5, kcal_y5,
    apoe4, egfr_y5,
    avesys_y5, hyper_y5, htnmed_y5, prevdm_y5, gluadj_y5,
    ldl_y5, hdl_y5, lipid_y5, crp_y5,
    kiddis_y5, chdyr5, stkyr5, estrgn_y5
  )

skewed <- c("kcal_y5", "crp_y5", "gluadj_y5")

table1 <- tbl_data %>%
  tbl_summary(
    by = sample,
    type = list(c(age_y5, avesys_y5) ~ "continuous"),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_of(skewed)    ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    value = list(
      c(htnmed_y5, prevdm_y5, lipid_y5, kiddis_y5, estrgn_y5, apoe4) ~ "Yes",
      gend01 ~ "Female"
    ),
    digits = list(all_continuous() ~ 1, all_categorical() ~ c(0, 1)),
    label = list(
      age_y5    ~ "Age, years",
      gend01    ~ "Female sex",
      race01    ~ "Race",
      grade01   ~ "Education",
      bmi_y5    ~ "BMI, kg/m²",
      waist_y5  ~ "Waist circumference, cm",
      smoke_y5  ~ "Smoking status",
      kcal_y5   ~ "Physical activity, kcal/week",
      avesys_y5 ~ "Systolic BP, mmHg",
      hyper_y5  ~ "Hypertension status",
      htnmed_y5 ~ "Antihypertensive medication",
      prevdm_y5 ~ "Diabetes",
      gluadj_y5 ~ "Glucose, mg/dL",
      ldl_y5    ~ "LDL cholesterol, mg/dL",
      hdl_y5    ~ "HDL cholesterol, mg/dL",
      lipid_y5  ~ "Lipid-lowering medication",
      crp_y5    ~ "CRP, mg/L",
      egfr_y5   ~ "eGFR, mL/min/1.73 m²",
      kiddis_y5 ~ "Self-reported kidney disease",
      chdyr5    ~ "Prevalent CHD",
      stkyr5    ~ "Prevalent stroke",
      estrgn_y5 ~ "Estrogen use",
      apoe4     ~ "APOE ε4 carrier"
    ),
    missing = "no"
  ) %>%
  modify_header(
    label ~ "**Characteristic**",
    all_stat_cols() ~ "**{level}**  \nN = {style_number(n)}"
  ) %>%
  modify_spanning_header(all_stat_cols() ~ "**Included in study for outcomes**") %>%
  modify_caption(paste0("**Table 1.** Baseline (CHS Year 5) characteristics of participants with klotho measured by ", measure_method, ", by analysis sample")) %>%
  modify_footnote(all_stat_cols() ~
                    "Mean (SD), median (IQR) or n (%). Participants included in more than one analysis appear in each relevant column. Total = participants in at least one analysis.") %>%
  bold_labels()

table1

table1 %>% as_gt() %>% gtsave(file.path(dir_desc, "table1_baseline.docx"))
table1 %>% as_gt() %>% gtsave(file.path(dir_desc, "table1_baseline.png"))
```

## Table 2: Characteristics by klotho tertile

Characteristics of all participants with klotho by tertile of klotho. Skewed variables (chosen from the histograms) are shown as median (IQR). Categorical variables are tested with χ² tests and continuous variables with the `gtsummary` default (Kruskal–Wallis).

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis

tbl_vars <- c(
  "age_y5", "gend01", "race01", "grade01", "smoke_y5",
  "egfr_y5", "kiddis_y5",
  "chdyr5", "stkyr5",
  "apoe4",
  "score134_y5", "digcor_y5"
)

tbl <- data_klot %>%
  left_join(data_clin, by = "idno") %>%
  mutate(tertile = ntile(klot_value, 3) %>% factor(labels = c("Low", "Mid", "High"))) %>%
  select(tertile, all_of(tbl_vars))

# Distribution of continuous variables
tbl %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_grid(~ Variable, scales = "free") +
  theme_minimal()

skewed <- c("age_y5", "score134_y5", "digcor_y5")

table2 <- tbl %>%
  tbl_summary(
    by = tertile,
    type = list(c(age_y5, egfr_y5, score134_y5, digcor_y5) ~ "continuous"),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_of(skewed)    ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    value = list(
      c(kiddis_y5, chdyr5, stkyr5, apoe4) ~ "Yes",
      gend01 ~ "Female"
    ),
    digits = list(all_continuous() ~ 1, all_categorical() ~ c(0, 1)),
    label = list(
      age_y5      ~ "Age, years",
      gend01      ~ "Female sex",
      race01      ~ "Race",
      grade01     ~ "Education",
      smoke_y5    ~ "Smoking status",
      chdyr5      ~ "Prevalent CHD",
      stkyr5      ~ "Prevalent stroke",
      apoe4       ~ "APOE ε4 carrier",
      score134_y5 ~ "3MSE score, Year 5",
      digcor_y5   ~ "DSST score, Year 5",
      egfr_y5     ~ "eGFR, mL/min/1.73 m²",
      kiddis_y5   ~ "Kidney disease (self-reported)"
    ),
    missing = "no"
  ) %>%
  add_p(test = list(all_categorical() ~ "chisq.test")) %>%
  modify_header(
    label ~ "**Characteristic**",
    all_stat_cols() ~ "**{level}**  \nN = {style_number(n)}"
  ) %>%
  modify_spanning_header(all_stat_cols() ~ paste0("**Tertiles of ", measure_method, "-measured Klotho**")) %>%
  bold_labels() %>%
  bold_p()

table2 %>% as_gt() %>% gtsave(file.path(dir_desc, "table2_klotho_tertiles.docx"))
table2 %>% as_gt() %>% gtsave(file.path(dir_desc, "table2_klotho_tertiles.png"))

print(table2)


```

Year 5 3MSE and DSST by klotho tertile. The overall Kruskal–Wallis p-value is shown in the panel title, with pairwise Wilcoxon tests above the boxes.

```{r}
plot_data <- tbl %>%
  filter(!is.na(tertile)) %>%
  pivot_longer(c(score134_y5, digcor_y5), names_to = "Test", values_to = "Score") %>%
  filter(!is.na(Score)) %>%
  mutate(Test = factor(Test, levels = c("score134_y5", "digcor_y5"), labels = c("3MSE", "DSST")))

facet_labs <- plot_data %>%
  group_by(Test) %>%
  summarise(p = kruskal.test(Score ~ tertile)$p.value) %>%
  mutate(lab = paste0(Test, "\nKruskal–Wallis p ",
                      if_else(p < 0.001, "< 0.001", paste("=", signif(p, 2)))))

plot_data <- plot_data %>% left_join(facet_labs %>% select(Test, lab), by = "Test")

tertile_cols <- c(Low = "#C6DBEF", Mid = "#6BAED6", High = "#2171B5")

p <- ggplot(plot_data, aes(x = tertile, y = Score, fill = tertile)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3, outlier.size = 0.8) +
  stat_compare_means(
    comparisons = list(c("Low", "Mid"), c("Mid", "High"), c("Low", "High")),
    method = "wilcox.test", label = "p.format", size = 3.2
  ) +
  scale_fill_manual(values = tertile_cols) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  facet_wrap(~ lab, scales = "free_y") +
  labs(x = paste0("Klotho tertile (", measure_method, ")"), y = "Score, Year 5") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"),
        panel.grid.major.x = element_blank())

p

ggsave(file.path(dir_desc, "boxplot_cognition_by_klotho_tertile.png"), p, height = 5, width = 7)

rm(plot_data, facet_labs, plot_data, tertile_cols, p)
```

## Klotho by APOE

### APOE ε4 carrier status

Klotho in APOE ε4 carriers vs non-carriers among participants in the Aim 2 populations, compared with a Welch t-test.

```{r}
plot_apoe <- data_clin %>%
  filter(idno %in% idnos_klot_aim2_dsst | idno %in% idnos_klot_aim2_3mse) %>%
  left_join(data_klot, by = "idno") %>%
  filter(!is.na(apoe4), !is.na(klot_value)) %>%
  mutate(apoe4 = factor(apoe4, levels = c("No", "Yes")))

p_apoe <- t.test(klot_value ~ apoe4, data = plot_apoe)$p.value
plot_apoe <- plot_apoe %>%
  mutate(lab = paste0(measure_method, " klotho\nT-test p ",
                      if_else(p_apoe < 0.001, "< 0.001", paste("=", signif(p_apoe, 2)))))

p <- ggplot(plot_apoe, aes(x = apoe4, y = klot_value, fill = apoe4)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3, outlier.size = 0.8) +
  stat_compare_means(
    comparisons = list(c("No", "Yes")),
    method = "t.test", label = "p.format", size = 3.2
  ) +
  scale_fill_manual(values = c(No = "#C6DBEF", Yes = "#2171B5")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  facet_wrap(~ lab) +
  labs(x = "APOE ε4 carrier", y = paste0("Klotho (", measure_method, "), Year 5")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"),
        panel.grid.major.x = element_blank())

p

ggsave(file.path(dir_desc, "boxplot_klotho_by_apoe4.png"), p, height = 5, width = 4)


```

Mean difference with 95% CI, from the t-test and the equivalent linear model.

```{r}
t.test(klot_value ~ apoe4, data = plot_apoe, conf.int = TRUE)   # Welch t-test, mean difference + 95% CI
lm(klot_value ~ apoe4, data = plot_apoe) |> tidy(conf.int = TRUE)  # mean difference (SD units)
```

Crude and adjusted (age, sex, race, eGFR) difference in klotho for carriers vs non-carriers.

```{r}
fit_crude_b <- lm(klot_value ~ apoe4, data = plot_apoe)
fit_adj_b   <- lm(klot_value ~ apoe4 + age_y5 + gend01 + race01 + egfr_y5, data = plot_apoe)

model_levels <- c("Crude", "Adjusted (age, sex, race, eGFR)")

ref_rows_b <- tibble(
  term = "apoe4No", estimate = 0, conf.low = NA_real_, conf.high = NA_real_,
  model = model_levels
)

est_b <- bind_rows(
  tidy(fit_crude_b, conf.int = TRUE) %>% mutate(model = model_levels[1]),
  tidy(fit_adj_b,   conf.int = TRUE) %>% mutate(model = model_levels[2])
) %>%
  filter(term == "apoe4Yes") %>%
  bind_rows(ref_rows_b) %>%
  mutate(carrier = factor(sub("apoe4", "", term), levels = c("No", "Yes")),
         model   = factor(model, levels = model_levels),
         is_ref  = carrier == "No")

n_lab_b <- plot_apoe %>%
  count(apoe4) %>%
  mutate(lab = if_else(apoe4 == "No",
                       paste0("Non-carrier (ref)\n(n = ", n, ")"),
                       paste0("ε4 carrier\n(n = ", n, ")")))

p <- ggplot(est_b, aes(x = estimate, y = carrier, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2,
                orientation = "y", position = position_dodge(width = 0.5), na.rm = TRUE) +
  geom_point(aes(shape = is_ref), size = 2.8, position = position_dodge(width = 0.5)) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 15), guide = "none") +
  scale_colour_manual(values = c("#9ECAE1", "#2171B5"), name = NULL) +
  scale_y_discrete(labels = setNames(n_lab_b$lab, n_lab_b$apoe4), limits = rev) +
  labs(x = "Difference in Klotho (SD) vs non-carriers, 95% CI", y = NULL,
       title = paste0(measure_method, " α-Klotho by APOE ε4 carrier status")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())

p

ggsave(file.path(dir_desc, "forest_klotho_by_apoe4.png"), p, height = 5, width = 8)
```

### APOE genotype

Klotho by APOE genotype. Overall Kruskal–Wallis test in the panel title, and pairwise Wilcoxon tests against ε3/ε3 above the boxes.

```{r}
plot_apoe_geno <- plot_apoe %>%
  filter(!is.na(apoecvd)) %>%
  mutate(apoe_genotype = factor(
    apoecvd,
    levels = c(22, 23, 33, 24, 34, 44),
    labels = c("ε2/ε2", "ε2/ε3", "ε3/ε3", "ε2/ε4", "ε3/ε4", "ε4/ε4")
  ))

p_geno <- kruskal.test(klot_value ~ apoe_genotype, data = plot_apoe_geno)$p.value

plot_apoe_geno <- plot_apoe_geno %>%
  mutate(lab = paste0(measure_method, " klotho\nKruskal-Wallis p ",
                      if_else(p_geno < 0.001, "< 0.001", paste("=", signif(p_geno, 2)))))

genotype_cols <- c(
  "ε2/ε2" = "#DEEBF7",
  "ε2/ε3" = "#C6DBEF",
  "ε3/ε3" = "#9ECAE1",
  "ε2/ε4" = "#6BAED6",
  "ε3/ε4" = "#2171B5",
  "ε4/ε4" = "#08306B"
)

p <- ggplot(plot_apoe_geno, aes(x = apoe_genotype, y = klot_value, fill = apoe_genotype)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3, outlier.size = 0.8) +
  stat_compare_means(
    comparisons = list(c("ε3/ε3", "ε2/ε3"),
                       c("ε3/ε3", "ε3/ε4"),
                       c("ε3/ε3", "ε4/ε4"),
                       c("ε3/ε3", "ε2/ε4"),
                       c("ε3/ε3", "ε2/ε2")),
    method = "wilcox.test", label = "p.format", size = 3.2
  ) +
  scale_fill_manual(values = genotype_cols) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  facet_wrap(~ lab) +
  labs(x = "APOE genotype", y = paste0("Klotho (", measure_method, "), Year 5")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"),
        panel.grid.major.x = element_blank())

p

ggsave(file.path(dir_desc, "boxplot_klotho_by_apoe_genotype.png"), p, height = 5, width = 10)
```

Carriers vs non-carriers among participants with known genotype.

```{r}
wilcox.test(klot_value ~ apoe4, data = plot_apoe_geno, conf.int = TRUE)   # Hodges–Lehmann shift + 95% CI
lm(klot_value ~ apoe4, data = plot_apoe_geno) |> tidy(conf.int = TRUE)    # mean difference (SD units)
```

## Klotho in prevalent dementia

Among participants in the cognition study, klotho in those with prevalent dementia at baseline vs everyone else, compared with a t-test.

```{r}
demclass_df <- data_clin %>%
  filter(!is.na(demclass)) %>%
  left_join(data_klot, by = "idno") %>%
  mutate(demclass_prevalent = if_else(demclass == "Prevalent", "Prevalent", "Non-prevalent"))

demclass_df %>%
  select(demclass_prevalent, klot_value) %>%
  tbl_summary(
    by = demclass_prevalent,
    type = klot_value ~ "continuous2",
    statistic = klot_value ~ c("{mean} ({sd})",
                               "{median} ({p25}, {p75})",
                               "{min}, {max}"),
    missing = "ifany"
  ) %>%
  add_p(test = klot_value ~ "t.test")

ggplot(demclass_df, aes(x = demclass_prevalent, y = klot_value, fill = demclass_prevalent)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +   # outliers are shown by the points
  geom_jitter(width = 0.15, alpha = 0.5) +
  stat_compare_means(
    comparisons = list(c("Non-prevalent", "Prevalent")),
    method = "t.test", label = "p.format", size = 3.2
  ) +
  labs(x = NULL, y = paste0("Klotho (", measure_method, "), Year 5")) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Aim 1: Klotho and incident dementia

## Analysis data

Klotho, covariates and follow-up for the Aim 1 population, with two outcomes: incident dementia, and incident dementia or MCI. Klotho tertiles are added for the survival curves.

```{r}
data_combined_aim1 <- full_join(data_klot, data_clin, by = "idno") %>%
  select(idno, klot_value, all_of(cov_m3), demclass, demtype, yearsfu) %>%
  filter(idno %in% idnos_klot_aim1) %>%
  mutate(
    tertile          = ntile(klot_value, 3),
    dem_incident     = case_when(demclass == "Incident" ~ TRUE, TRUE ~ FALSE),
    dem_incident_MCI = case_when(demclass %in% c("MCI", "Incident") ~ TRUE, TRUE ~ FALSE)
  )
```

## Survival curves

Unadjusted curves by klotho tertile with risk tables and log-rank p-values.

| `type`       | Transformation   |
|:-------------|:-----------------|
| `"survival"` | $x$              |
| `"risk"`     | $1 - x$          |
| `"cumhaz"`   | $-\log(x)$       |
| `"cloglog"`  | $\log(-\log(x))$ |

| Estimator | What it estimates | In R |
|:-----------------------|:-----------------------|:-----------------------|
| **Kaplan–Meier** | Survival $S(t)$, or $1 - S(t)$ for event risk | `survfit(..., stype = 1, ctype = 1)` |
| **Nelson–Aalen** | Cumulative hazard $H(t)$ | `survfit(..., stype = 2, ctype = 1)` gives survival as $\exp(-H)$ |
| **Aalen–Johansen** | Cumulative incidence with competing risks | `survfit()` with a multi-state formulation |

Dementia-free survival:

```{r}
survfit2(Surv(yearsfu, dem_incident) ~ tertile, data = data_combined_aim1) %>%
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable() +
  add_pvalue(location = "annotation") +     # log-rank p
  scale_color_manual(values = c("#C6DBEF", "#6BAED6", "#2171B5")) +
  scale_fill_manual(values  = c("#C6DBEF", "#6BAED6", "#2171B5"))
```

Cumulative incidence of dementia or MCI:

```{r}
survfit2(Surv(yearsfu, dem_incident_MCI) ~ tertile, data = data_combined_aim1) %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable() +
  add_pvalue(location = "annotation") +     # log-rank p
  scale_color_manual(values = c("#C6DBEF", "#6BAED6", "#2171B5")) +
  scale_fill_manual(values  = c("#C6DBEF", "#6BAED6", "#2171B5")) +
  labs(x = "Years since Year 5", y = "Cumulative incidence of dementia")
```

## Cox models

Outcome: incident dementia. Results are saved to `02_cox_dementia/dem_incident/<model>/`.

### Model 0 (crude)

```{r}
res_m0 <- run_cox("dem_incident", cov_m0)
```

### Model 1

```{r}
res_m1 <- run_cox("dem_incident", cov_m1)
save_forest_cox(res_m1)
```

### Model 1, stratified by APOE ε4

```{r}
res_m1_apoe4 <- run_cox("dem_incident", setdiff(cov_m1, "apoe4"),
                        data = data_combined_aim1 %>% filter(apoe4 == "Yes"),
                        model_name = "cov_m1_apoe4_carriers")
res_m1_nonapoe4 <- run_cox("dem_incident", setdiff(cov_m1, "apoe4"),
                           data = data_combined_aim1 %>% filter(apoe4 == "No"),
                           model_name = "cov_m1_apoe4_noncarriers")

p_yes <- res_m1_apoe4$results    %>% format_forest() %>% plot_forest_hr()
p_no  <- res_m1_nonapoe4$results %>% format_forest() %>% plot_forest_hr()

p <- (p_yes + ggtitle("APOE ε4 positive")) + (p_no + ggtitle("APOE ε4 negative"))
p

ggsave(file.path(dir_cox, "dem_incident", "forest_cov_m1_by_apoe4.png"),
       p, height = 5, width = 15, bg = "white")
```

### Model 1, stratified by sex

Participants of "Other" race are excluded.

```{r}
res_m1_fem <- run_cox("dem_incident", setdiff(cov_m1, "gend01"),
                      data = data_combined_aim1 %>% filter(gend01 == "Female" & race01 != "Other"),
                      model_name = "cov_m1_women")
res_m1_mal <- run_cox("dem_incident", setdiff(cov_m1, "gend01"),
                      data = data_combined_aim1 %>% filter(gend01 == "Male" & race01 != "Other"),
                      model_name = "cov_m1_men")

p_fem <- res_m1_fem$results %>% format_forest(drop_terms = "race01Other") %>% plot_forest_hr()
p_mal <- res_m1_mal$results %>% format_forest(drop_terms = "race01Other") %>% plot_forest_hr()

p <- (p_fem + ggtitle("Female")) + (p_mal + ggtitle("Male"))
p

ggsave(file.path(dir_cox, "dem_incident", "forest_cov_m1_by_sex.png"),
       p, height = 5, width = 15, bg = "white")
```

### Model 2

```{r}
# res_m2 <- run_cox("dem_incident", cov_m2)
# save_forest_cox(res_m2, height = 5)
```

### Model 3 (sensitivity)

```{r}
# res_m3 <- run_cox("dem_incident", cov_m3)
# save_forest_cox(res_m3, height = 7)
```

# Aim 2: Klotho and cognitive trajectories

## Analysis data

Long format with one row per participant and study year, with 3MSE (`score134`) and DSST (`digcor`) scores and klotho.

```{r}
data_clin_long <- data_clin %>%
  left_join(data_klot, by = "idno") %>%
  pivot_longer(
    cols            = matches("^(score134|digcor)_y\\d+$"),
    names_to        = c(".value", "year"),
    names_pattern   = "(score134|digcor)_y(\\d+)",
    names_transform = list(year = as.integer)
  )
```

Models included in the combined forest plots (Model 3 is left out).

```{r}
rows_lmm <- tribble(
  ~model,                 ~label,         ~group,
  "m0",                   "Crude",        "Overall",
  "m1",                   "Model 1",      "Overall",
  "m2",                   "Model 2",      "Overall",
  # "m3",                 "Model 3",      "Overall",
  "m1_apoe4_carriers",    "ε4 carriers",  "Stratified by APOE ε4 (Model 1)",
  "m1_apoe4_noncarriers", "Non-carriers", "Stratified by APOE ε4 (Model 1)"
)
```

## 3MSE

```{r}
res_3mse <- run_lmm_models("score134", idnos_klot_aim2_3mse, test_label = "3MSE")
```

```{r}
res_3mse$m3$results
```

Klotho estimates across models: difference in yearly change (klotho × time) and difference at Year 5.

```{r}
p_slope <- plot_forest_lmm(res_3mse[rows_lmm$model], rows_lmm, term = "klot_value:time",
                           xlab = "Difference in yearly 3MSE change per SD Klotho (points)")
p_level <- plot_forest_lmm(res_3mse[rows_lmm$model], rows_lmm, term = "klot_value",
                           xlab = "Difference in 3MSE at year 5 per SD Klotho (points)")
p_slope
p_level

ggsave(file.path(dir_lmm, "forest_3mse_slope.png"),
       p_slope, width = 10, height = 3.8, dpi = 300, bg = "white")
```

## DSST

```{r}
res_dsst <- run_lmm_models("digcor", idnos_klot_aim2_dsst, test_label = "DSST")
```

```{r}
p_slope <- plot_forest_lmm(res_dsst[rows_lmm$model], rows_lmm, term = "klot_value:time",
                           xlab = "Difference in yearly DSST change per SD Klotho (points)")
p_level <- plot_forest_lmm(res_dsst[rows_lmm$model], rows_lmm, term = "klot_value",
                           xlab = "Difference in DSST at year 5 per SD Klotho (points)")
p_slope
p_level

ggsave(file.path(dir_lmm, "forest_dsst_slope.png"),
       p_slope, width = 10, height = 3.8, dpi = 300, bg = "white")
```

# Aim 3: Genetics

## Data

Genotypes for selected *KL* SNPs (number of effect alleles: 0, 1 or 2).

```{r}
data_genetics <- read_csv(
  file.path(data_path, "Meineche_10237", "Klotho SNPs", "CHSCC_TOPMedWGS_July2026_snps.csv")
) %>%
  mutate(idno = as.character(idno))
```

rs9527025 and rs9536314 are in near-perfect linkage disequilibrium, as expected since they are inherited together as the KL-VS haplotype:

```{r}
cor(data_genetics$rs9527025_G, data_genetics$rs9536314_T)
```

SNP annotation (gene names) from g:Profiler g:SNPense.

```{r}
data_snp_info <- read.csv(
  file.path(data_path, "Meineche_10237", "Klotho SNPs", "gProfiler_gSNPense_2026-10-05_12-14-26.csv"),
  skip = 1
) %>%
  rename(SNP = id)
```

## Klotho levels by SNP

### Boxplots

Klotho by number of effect alleles for each SNP, with pairwise t-tests.

```{r}
data_genetics_klot <- left_join(data_genetics, data_klot, by = "idno") %>%
  filter(!is.na(klot_value)) %>%
  pivot_longer(cols = -c(idno, klot_value), names_to = "SNP", values_to = "Mutations") %>%
  mutate(Mutations = factor(Mutations))

p <- data_genetics_klot %>%
  ggplot(aes(x = Mutations, y = klot_value)) +
  geom_boxplot(aes(fill = Mutations)) +
  facet_wrap(~ SNP, ncol = 4) +
  stat_compare_means(
    comparisons = list(c("0", "1"), c("1", "2"), c("0", "2")),
    method = "t.test", label = "p.format", size = 3.2
  ) +
  labs(y = paste0("Klotho SD (", measure_method, ")")) +
  theme_minimal()

p

ggsave(file.path(dir_gen, "boxplot_klotho_by_snp.png"), p, height = 8, width = 12)
```

### Linear models

For each SNP, `klot_value ~ genotype` with 0 copies as reference. The per-genotype p-value compares 1 or 2 copies with 0 copies; the overall F-test (panel title) tests any difference between genotypes.

```{r}
geno_cols <- c("0" = "#9ECAE1", "1" = "#2171B5", "2" = "#08306B")

data_genetics_klot <- data_genetics %>% 
  pivot_longer(cols = -c(idno), names_to = 'SNP', values_to = 'Mutations') %>%
  left_join(data_klot, by = 'idno')

# One model per SNP
model_df <- data_genetics_klot %>%
  mutate(Mutations = factor(Mutations, levels = c(0, 1, 2))) %>%
  nest(.by = SNP) %>%
  mutate(
    fit       = map(data, \(d) lm(klot_value ~ Mutations, data = d)),
    coef      = map(fit,  \(f) tidy(f, conf.int = TRUE)),
    overall_p = map_dbl(fit, \(f) glance(f)$p.value)
  )

# n per SNP and genotype
n_tab <- data_genetics_klot %>%
  filter(!is.na(klot_value), !is.na(Mutations)) %>%
  count(SNP, Level = as.character(Mutations), name = "n")

# Results table
results_snp <- model_df %>%
  select(SNP, overall_p, coef) %>%
  unnest(coef) %>%
  filter(term != "(Intercept)") %>%
  mutate(Level = sub("Mutations", "", term)) %>%
  bind_rows(model_df %>% transmute(SNP, overall_p, Level = "0", estimate = 0)) %>%
  left_join(n_tab, by = c("SNP", "Level")) %>%
  mutate(
    Level  = factor(Level, levels = c(0, 1, 2)),
    is_ref = Level == "0",
    lab    = if_else(
      is_ref,
      paste0("Reference   n = ", n),
      sprintf("%.2f (%.2f to %.2f)   p %s   n = %d",
              estimate, conf.low, conf.high, fmt_p(p.value), n)
    )
  ) %>%
  rename(SNP_allele = SNP) %>%
  separate(SNP_allele, sep = "_", into = c("SNP", "allele"), remove = FALSE) %>%
  left_join(data_snp_info, by = "SNP") %>%
  mutate(facet_lab = paste0(SNP_allele, " (", gene_names, ")",
                            "\nOverall F-test p ", fmt_p(overall_p))) %>%
  arrange(SNP, Level)

# Forest plot
x_text <- max(results_snp$conf.high, na.rm = TRUE) * 1.1   # where the text column starts

p <- ggplot(results_snp, aes(x = estimate, y = Level, colour = Level)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2,
                orientation = "y", linewidth = 0.8, na.rm = TRUE) +
  geom_point(aes(shape = is_ref), size = 3) +
  geom_text(aes(x = x_text, label = lab), hjust = 0, size = 3.2, colour = "grey20") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 15), guide = "none") +
  scale_colour_manual(values = geno_cols, guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 1.2))) +   # room for the text
  facet_wrap(~ facet_lab, ncol = 2) +
  labs(x = "Difference in Klotho (SD) vs 0 copies, 95% CI",
       y = "Copies",
       title = paste0(measure_method, " α-Klotho by genotype")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())

p

ggsave(file.path(dir_gen, "forest_klotho_by_snp.png"), p,
       height = 2 + 1.6 * ceiling(n_distinct(results_snp$SNP) / 2), width = 14, bg = "white")
```

SNPs included:

```{r}
cat(unique(results_snp$SNP), sep = "\n")
```

## KL-VS

Two *KL* variants, rs9536314 (p.F352V) and rs9527025 (p.C370S), are in strong linkage disequilibrium and segregate together as a functional haplotype, KL-VS. Carrying one copy, but not two, of the KL-VS haplotype (KL-VS heterozygosity, KL-VS^het^) has been linked to higher circulating klotho ([Driscoll et al., PLoS One 2022](https://doi.org/10.1371/journal.pone.0267298)).

Because the two variants are in near-perfect LD, KL-VS status is defined from rs9527025 alone: heterozygous (1 copy) = `Het+`, otherwise (0 or 2 copies) = `Het-`.

```{r}
data_klvs <- data_genetics %>%
  mutate(KL_VS = factor(case_when(rs9527025_G == 1 ~ "Het+", TRUE ~ "Het-"))) %>%
  left_join(data_klot, by = "idno") %>%
  left_join(data_clin,  by = "idno")

table(data_klvs$KL_VS, useNA = "ifany")
```

### Klotho levels

```{r}
p <- data_klvs %>%
  filter(!is.na(klot_value), !is.na(KL_VS)) %>%
  ggplot(aes(x = KL_VS, y = klot_value, fill = KL_VS)) +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3, outlier.size = 0.8) +
  stat_compare_means(
    comparisons = list(c("Het-", "Het+")),
    method = "t.test", label = "p.format", size = 3.2
  ) +
  scale_fill_manual(values = c("Het-" = "#C6DBEF", "Het+" = "#2171B5")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +   # room for the p-value
  labs(x = NULL, y = paste0("Klotho SD (", measure_method, ")")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank())

p

ggsave(file.path(dir_klvs, "boxplot_klotho_by_klvs.png"), p, height = 8, width = 14, bg = "white")
```

Difference in klotho for KL-VS Het+ vs Het-, crude and across covariate sets.

```{r}
cov_m05 = c('age_y5', 'gend01')

model_sets <- list(
  "Crude"     = character(0),
  "Model 0.5" = cov_m05,
  "Model 1"   = cov_m1,
  "Model 2"   = cov_m2,
  "Model 3"   = cov_m3
)

model_cols <- c("#C6DBEF", "#9ECAE1", "#6BAED6", "#2171B5", "#08306B")

fits_klvs <- map(model_sets, \(cv)
  lm(reformulate(c("KL_VS", cv), response = "klot_value"), data = data_klvs)
)

est_klvs <- imap_dfr(fits_klvs, \(f, nm) {
  mf <- model.frame(f)                       # the rows actually used in the model
  tidy(f, conf.int = TRUE) %>%
    filter(str_starts(term, "KL_VS")) %>%
    mutate(model     = nm,
           n         = nrow(mf),
           n_carrier = sum(mf$KL_VS == "Het+"))
}) %>%
  mutate(
    model = factor(model, levels = names(model_sets)),
    lab   = sprintf("%.2f (%.2f to %.2f)   p %s   n = %d (%d Het+)",
                    estimate, conf.low, conf.high, fmt_p(p.value), n, n_carrier)
  )

est_klvs

write_xlsx(est_klvs, file.path(dir_klvs, "lm_klotho_by_klvs.xlsx"))

# Covariates in each model
adj_text <- imap_chr(model_sets, \(cv, nm)
  paste0(nm, ": ", if (length(cv) == 0) "unadjusted" else paste(nice_cov(cv), collapse = ", "))
)
cat(adj_text, sep = "\n")
writeLines(adj_text, file.path(dir_klvs, "lm_klotho_by_klvs_adjustment.txt"))

# Forest plot
x_text <- max(est_klvs$conf.high) + 0.1 * diff(range(c(est_klvs$conf.low, est_klvs$conf.high)))

p <- ggplot(est_klvs, aes(x = estimate, y = model, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2,
                orientation = "y", linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(aes(x = x_text, label = lab), hjust = 0, size = 3.2, colour = "grey20") +
  scale_colour_manual(values = model_cols, guide = "none") +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 1.3))) +   # room for the text
  labs(x = "Difference in Klotho (SD), KL-VS Het+ vs Het-, 95% CI", y = NULL,
       title = paste0(measure_method, " α-Klotho by KL-VS carrier status")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())

p

ggsave(file.path(dir_klvs, "forest_lm_klotho_by_klvs.png"), p, height = 4, width = 11, bg = "white")
```

### KL-VS and incident dementia

Cox models of incident dementia (and dementia or MCI) on KL-VS status, excluding participants of "Other" race. Results are saved to `04_genetics/klvs/cox/<outcome>/<model>/`.

```{r}
dir_klvs_cox <- file.path(dir_klvs, "cox")

data_klvs_cox <- data_klvs %>%
  mutate(
    dem_incident     = case_when(demclass == "Incident" ~ TRUE, TRUE ~ FALSE),
    dem_incident_MCI = case_when(demclass %in% c("MCI", "Incident") ~ TRUE, TRUE ~ FALSE)
  ) %>%
  filter(race01 != "Other")
```

#### Incident dementia

```{r}
res_klvs_dem <- map(c("cov_m05", "cov_m1", "cov_m2"), \(adj) {
  res <- run_cox("dem_incident", get(adj),
                 data = data_klvs_cox, variable_of_interest = "KL_VS",
                 model_name = adj, out_dir = dir_klvs_cox)
  print(save_forest_cox(res, drop_terms = "race01Other"))
  res
}) %>%
  set_names(c("cov_m05", "cov_m1", "cov_m2"))
```

#### Incident dementia or MCI

```{r}

res_klvs_dem <- map(c("cov_m05", "cov_m1", "cov_m2"), \(adj) {
  res <- run_cox("dem_incident_MCI", get(adj),
                 data = data_klvs_cox, variable_of_interest = "KL_VS",
                 model_name = glue('{adj}_MCI'), out_dir = dir_klvs_cox)
  print(save_forest_cox(res, drop_terms = "race01Other"))
  res
}) %>%
  set_names(c("cov_m05", "cov_m1", "cov_m2"))


```
